---
title: "Statistical analysis of trench data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Statistical analysis of trench data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  rmarkdown.html_vignette.check_title = FALSE,
  comment = "#>"
)

library(TrenchR)
library(dplyr)
library(magrittr)
library(knitr)
```

## Overview

This vignette introduces the `TrenchR` functions which can be used to
statistically analyse trench datasets, and shows how they can be used to
reproduce the statistical summary table in Münch et al. (2017).

## Trench summary

`summarizeTrench()` retrieves information on the number and distances of sampled
trench profiles, and on statistical data of a requested trench variable.

Per default, information on the sampled profiles are returned:

```{r}
summarizeTrench(t13.trench1)
```

If you supply a trench variable, `summary()`-like information on that variable
will be included:

```{r}
summarizeTrench(t13.trench1, .var = "d18O")
```

## Trench decorrelation scales and effective degrees of freedom

An insightful analysis for a trench dataset is to analyse the decorrelation
length of the data in horizontal and/or vertical direction. For this, you can
use the function `estimateTrenchDecorrelation()`, which can handle both
equidistantly and non-equidistantly sampled data (see the function's
documentation for details on the underlying estimation methods):

```{r}
estimateTrenchDecorrelation(t15.trench2, .var = "d18O", vscale = "depth")
```

The decorrelation length in horizontal direction is related to the question how
well neighbouring trench profiles are correlated; for this see also the section
[Estimating trench correlations and signal-to-noise ratios] below.

With a decorrelation length estimate in a given direction, you can use
`getEffectiveTrenchDOF()` to calculate the effective degrees of freedom (DOF) of
your trench dataset in that direction, e.g.

```{r}
getEffectiveTrenchDOF(lambda = 1.5, N = 10, delta = 2)
```

which is equivalent to

```{r}
getEffectiveTrenchDOF(lambda = 1.5, positions = seq(0, 18, 2))
```

## Trench variance

The variances of the horizontal and vertical data variations in a trench can be
assessed with

```{r}
estimateTrenchVariance(t13.trench2, .var = "d18O")
```

including confidence intervals based on the Chi-Squared distribution and the
DOF in both trench directions. Default for the latter are the number of data
points in the trench directions; to account for autocorrelation of the data the
DOF can be supplied also manually (see also `getEffectiveTrenchDOF()` above).

## Estimating trench correlations and signal-to-noise ratios

An interesting application for a trench dataset is to study how the correlation
between neighbouring profiles varies with their separation distance. In
`TrenchR`, you can assess this question using
`estimateInterProfileCorrelation()`, which calculates the average pairwise
correlation between trench profiles for given separation distances (including
some tolerance such that the inter-profile distances do not need to exactly
match the asked distances):

```{r}
estimateInterProfileCorrelation(t13.trench2, distances = c(10, 30))
```

As you can see, the output gives the number of found profile pairs for each
distance, their average correlation, and the associated standard deviation and
standard error of the mean estimate (see `?estimateInterProfileCorrelation()`
for details and more information on the error estimation).

This inter-profile correlation can be used as a measure for the signal-to-noise
ratio (SNR) of your trench dataset:

```{r}
estimateSNR(t13.trench2, distances = c(10, 30))
```

which is based on `estimateInterProfileCorrelation()` and is calculated from the
mean across the average pairwise profile correlations evaluated at the given
profile separation distances via $\mathrm{SNR} = \overline{c}\left(1 -
\overline{c}\right)^{-1}$, where $\overline{c}$ is the mean correlation.

To calculate the correlation between the individual profiles from two separate
trenches (assuming the two trenches were sampled at the same time) you can use
the function `estimateInterTrenchCorrelation()`:

```{r}
raw <- estimateInterTrenchCorrelation(t13.trench1, t13.trench2)
opt <- estimateInterTrenchCorrelation(t13.trench1, t13.trench2,
                                      optimize = TRUE, lag = -4 : 4)

str(raw)
str(opt)

mean(raw)
mean(opt$cor)
```

The first result is simply a matrix of the correlations for all possible pairs
of profiles between the two trenches. In the second example, some vertical
shifts of $\pm 4$ (depth) bins are allowed via the `lag` parameter in order to
optimise (i.e. maximise) the correlations; the result in this case is a list of
two matrices where the first matrix includes the optimised correlations between
the profiles and the second matrix the lags for which the optimal correlations
were observed.

In addition to the individual profiles, you can also directly estimate the
correlation between two trenches based on their mean profiles,

```{r}
getMeanProfileCorrelation(t13.trench1, t13.trench2, lag = -4 : 4)
```

which also returns the corresponding root-mean-square deviation
(RMSD). Similarly to `estimateInterTrenchCorrelation()`, the `lag` parameter
allows vertical shifts of the mean profiles relative to each other, and the
function output displays the optimal correlation (and RMSD) across the given
lags, where the `vscale_shift` gives the optimal lag in physical units, which is
based on the `res` function parameter. This parameter per default uses the
inherent resolution of the trench datasets, but you can specify a different
resolution which the mean profiles are interpolated on prior to analysing the
optimal correlation (note that you need to adjust the `lag` parameter in that
case since that is measured relative to the resolution).

## Reproducing Münch et al., 2017, Table 2

In the following we use the above functions to reproduce the statistical results
displayed in Table 2 of Münch et al. (2017) for comparing the `t13.trench1`,
`t13.trench2`, `t15.trench2`, and `t15.trench2` datasets.

We first define a wrapper function that collects all relevant results for a
given trench dataaset in a single data frame:

```{r fun_dev}
#' Wrapper function to obtain statistical trench properties
#' 
#' @param x a trench dataset.
#' @param lambda.h horizontal decorrelation length for dataset in `x`.
#' @param lambda.v vertical decorrelation length for dataset in `x`.
#' @param distances a vector of inter-profile distances for estimating the SNR.
#' 
getTrenchStats <- function(x, lambda.h, lambda.v, distances) {

  # trench name from calling function
  nm <- deparse(substitute(x))

  # trench summary information
  s <- summarizeTrench(x, .var = "d18O")

  # effective degrees of freedom of trench
  dof.h <- getEffectiveTrenchDOF(lambda = lambda.h,
                                 positions = getX(x)) %>%
    round(1)
  dof.v <- getEffectiveTrenchDOF(lambda = lambda.v,
                                 positions = getZ(x)) %>%
    round(1)

  # trench variances for d18O
  d18O_var <- estimateTrenchVariance(x, .var = "d18O",
                                     dof.h = dof.h,
                                     dof.v = dof.v)

  # signal-to-noise ratio for d18O
  snr <- estimateSNR(x, distances, lambda = lambda.h)

  trench_stats <- data.frame(

    nprof = round(s$Nprofiles),
    d_min = round(s$d_min, 1),
    d_max = round(s$d_max, 1),
    d18O_min = round(s$min),
    d18O_max = round(s$max),
    d18O_mean = round(s$mean, 1),
    d18O_sd = round(s$sd, 1),
    d18O_var_h = round(d18O_var$var[1], 1),
    d18O_var_h_lower = round(d18O_var$lower[1], 1),
    d18O_var_h_upper = round(d18O_var$upper[1], 1),
    d18O_var_v = round(d18O_var$var[2], 1),
    d18O_var_v_lower = round(d18O_var$lower[2], 1),
    d18O_var_v_upper = round(d18O_var$upper[2], 1),
    snr = round(snr$snr, 1),
    snr_se = round(snr$se, 1)
  )

  trench_stats %>%
    t() %>%
    magrittr::set_colnames(nm)

}
```

We define the input data as used in the paper, estimate the average
decorrelation lengths, and then collect the results and print them as a table:

```{r stats}
# use data as in paper
T13.1 <- dplyr::filter(t13.trench1, profileName != "T13-1-01")
T13.2 <- t13.trench2
T15.1 <- dplyr::filter(t15.trench1, depth <= 175.5, profileName != "T15-1-DUNE1")
T15.2 <- dplyr::filter(t15.trench2, depth <= 175.5)

# estimate decorrelation length of trench data
# for horizontal direction, use densely sampled T13-1 trench
lambda.h <- estimateTrenchDecorrelation(T13.1, "horizontal")
# for vertical direction, average results for each trench season
lambda.v.t13 <- mean(estimateTrenchDecorrelation(T13.1, "vertical"),
                     estimateTrenchDecorrelation(T13.2, "vertical"))
lambda.v.t15 <- mean(estimateTrenchDecorrelation(T15.1, "vertical"),
                     estimateTrenchDecorrelation(T15.2, "vertical"))

# get statistical results for all four trenches
table_results <- cbind(
  getTrenchStats(T13.1, lambda.h, lambda.v.t13, distances = seq(10, 35, 5)),
  getTrenchStats(T13.2, lambda.h, lambda.v.t13, distances = c(10, 20, 30)),
  getTrenchStats(T15.1, lambda.h, lambda.v.t15, distances = seq(10, 35, 5)),
  getTrenchStats(T15.2, lambda.h, lambda.v.t15, distances = seq(10, 35, 5))
)

# print as a nice table
knitr::kable(table_results)
```

We note that some of the error limits differ slightly from the corresponding
values in the published table due to rounding differences and updates to the
error estimation code.

Finally, we obtain the optimal correlation and RMSD between the two trenches
from each season:

```{r}
# get mean profile opimization and print as table
c1 <- getMeanProfileCorrelation(T13.1, T13.2, lag = -4 : 4)
c2 <- getMeanProfileCorrelation(T15.1, T15.2, res = 0.5, lag = -12 : 12)

data.frame(shift = round(c(c1$vscale_shift, c2$vscale_shift), 2),
           correlation = round(c(c1$correlation, c2$correlation), 2),
           rmsd = round(c(c1$rmsd, c2$rmsd), 2)) %>%
  t() %>%
  magrittr::set_colnames(c("T13-1 vs T13-2", "T15-1 vs T15-2")) %>%
  knitr::kable()

```

## References

Münch, T., Kipfstuhl, S., Freitag, J., Meyer, H., and Laepple, T.: Constraints
on post-depositional isotope modifications in East Antarctic firn from analysing
temporal changes of isotope profiles, The Cryosphere, 11(5), 2175-2188, doi:
[10.5194/tc-11-2175-2017](https://doi.org/10.5194/tc-11-2175-2017), 2017.