---
title: "Introduction to TrenchR"
author: "Thomas M端nch"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to TrenchR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  rmarkdown.html_vignette.check_title = FALSE,
  comment = "#>"
)

library(TrenchR)
library(tibble)
library(dplyr)

```

## What is a trench?

A "trench" denotes a spatial array of profiles that tabulate climate proxy data
over depth or time, with the profiles being sampled longitudinally at regular
or irregular intervals. The R package `TrenchR` implements routines that are
targeted at the analysis and the plotting of the spatial (longitudinal) and
vertical (temporal) variations of the climate proxy data derived from such
trenches.


## Generic `TrenchR` dataset structure

In order to optimally work with `TrenchR` and facilitate the application of
`TrenchR` functions, it is convenient to prepare your "trench" dataset in a way
that adheres to the `TrenchR` generic dataset structure. In the following, this
generic structure is introduced and it is shown in a small example how to shape
your own data into this structure.

### The structure

`TrenchR` comes with four trench datasets: `t13.trench1`, `t13.trench2`,
`t15.trench1`, and `t15.trench2` (M端nch et
al. [2016](https://doi.org/10.5194/cp-12-1565-2016),
[2017](https://doi.org/10.5194/tc-11-2175-2017)). Since `t13.trench2` is the
smallest of these four, it is easiest to have a look at this dataset to
understand the generic structure:

```{r}
t13.trench2
```

You can see that the generic dataset structure is nothing more than a `tibble`
(it can also be a simple `data.frame`), however, with the individual proxy
profiles bound consecutively by row and with a fixed nomenclature applied to the
column names.

The minimum requirement is a tibble with columns `profileName` and
`sampleNumber` plus some actual data column. Here, the column `profileName`
indicates the position of the profile within the trench via its name, while the
column `sampleNumber` labels the individual proxy data samples within a profile:

```{r}
nrow(t13.trench2)

unique(t13.trench2$profileName)
unique(t13.trench2$sampleNumber)
```

thus, the trench `t13.trench2` contains four profiles labelled "T13-2-01",
"T13-2-02", "T13-2-03", and "T13-2-04", and each profile includes 38 proxy
samples.

In addition, you can provide information on the horizontal profile positions
within the trench in a column named `profilePosition`, and, if applicable, on
the top heights of the profiles relative to some horizontal reference in the
column `surfaceHeight`. While these additional two columns are not mandatory in
general, they are required for some `TrenchR` functions.

Usually, there is also some vertical scale associated with the samples and
sample numbers. This can be "depth", as for `t13.trench2`, or "time", or any
other applicable scaling. `TrenchR` functions, which need to use this vertical
scale in order to work, include the function parameter `vscale` to specify the
name of the vertical scale column. In this context it is also important to note
that for `TrenchR` we **inherently assume** that this vertical scale is
_absolute_, i.e. that it is the same for all profiles.

### Creating a trench dataset

Suppose you have sampled a trench with three profiles, where each profile
includes two samples. Then, one easy way to shape these data into the generic
structure is to create a data frame for each profile and then bind these data
frames together by row.

Step by step this means:

```{r}
# create a data frame for each profile
prf1 <- tibble::tibble(
  sampleNumber = 1 : 2, depth = c(2, 4), my_data = c(1.1, 2.1))
prf2 <- tibble::tibble(
  sampleNumber = 1 : 2, depth = c(2, 4), my_data = c(2.2, 1.2))
prf3 <- tibble::tibble(
  sampleNumber = 1 : 2, depth = c(2, 4), my_data = c(0.3, 4.3))

# bind them together
my_trench <- dplyr::bind_rows(profile1 = prf1, profile2 = prf2, profile3 = prf3,
                              .id = "profileName")

my_trench
```

Since profile positions and/or surface heights are a single value for each
profile, you can directly provide these information in the first step:

```{r}
prf1 <- dplyr::mutate(prf1, profilePosition = 0, surfaceHeight = 2)
prf2 <- dplyr::mutate(prf2, profilePosition = 5, surfaceHeight = 0)
prf3 <- dplyr::mutate(prf3, profilePosition = 13, surfaceHeight = 1)

my_trench <- dplyr::bind_rows(profile1 = prf1, profile2 = prf2, profile3 = prf3,
                              .id = "profileName")

my_trench
```

And now you are already good to go to apply all the nice `TrenchR` functions to
your trench dataset!

## Applying `TrenchR`

### Extracting trench information

For quick access, `TrenchR` provides functions to directly retrieve information
on the trench dataset, such as the horizontal profile positions,

```{r}
getX(t13.trench2)
```

the vertical scale,

```{r}
getZ(t13.trench2) # default is the same as getZ(t13.trench2, .var = "depth")
```

and the trench surface profile:

```{r}
getSurfaceProfile(t13.trench2)
```

For such undulating trench surfaces, `getFirstCompleteDepthBin` extracts the
value of the first bin of the vertical scale dimension (e.g. "depth") for which
a complete dataset across all trench profiles is available:

```{r}
getFirstCompleteDepthBin(t13.trench2)
getFirstCompleteDepthBin(t15.trench1)
```

If you already might have an estimate of the horizontal autocorrelation ($a_1$)
of your trench data at hand, you can use `getEffectiveTrenchDOF` to calculate
the effective horizontal degrees of freedom of your trench dataset:

```{r}
getEffectiveTrenchDOF(a1 = 0.5, N = 10, delta = 2)
# is equivalent to:
getEffectiveTrenchDOF(a1 = 0.5, positions = seq(0, 18, 2))
```

### Two-dimensional dataset

The inherent two-dimensional structure of a trench dataset can be directly
visualised in matrix form; while this matrix representation is currently not
needed for using any of the below `TrenchR` functions, since they all accept the
generic dataset structure as input, it might be handy for external functions
and usage.

```{r}
make2D(t13.trench2)

# Note: use simplify = TRUE to really have the output class `matrix`
```

The default is to extract the data column labelled `d18O`; to extract any other
data column, use

```{r}
make2D(t13.trench2, .var = "dxs") # insert your data column name for `.var`
```

### Trench  mean

The mean profile from averaging in horizontal direction across all available
profiles can be obtained directly from

```{r}
makeMean(t13.trench2)
```

As you can see, the default method is to output the mean profile as a data frame
including the vertical scale column (here, "depth"); adjust the `vscale`
argument if your dataset uses another name for the vertical scale
column.

You can also output the mean profile as a normal vector:

```{r}
makeMean(t13.trench2, df = FALSE)
```

Analogously to R's `mean` function, you can control the handling of missing
values,

```{r}
makeMean(t13.trench2, df = FALSE, na.rm = TRUE)
```

and similarly to `make2D`, the function parameter `.var` controls for which data
variable the mean profile is calculated.

### Plotting a two-dimensional visualisation

The two-dimensional proxy data variations can be directly visualised using the
function `plot2D`, either as a filled image plot,

```{r, fig.height = 5.25, fig.width = 7}
plot2D(t13.trench1, xlim = c(-0.1, 46), ylim = c(114, 0), zlim = c(-55, -35),
       ylab = "Depth (cm)", hadj = 0.2, line.v = 3)
```

or as a filled contour plot:

```{r, fig.height = 5.25, fig.width = 7}
plot2D(t13.trench1, xlim = c(-0.1, 46), ylim = c(114, 0), zlim = c(-55, -35),
       ylab = "Depth (cm)", hadj = 0.02, line.v = 3,
       filledContour = TRUE, fill = TRUE)
```

There are a couple more options to adjust and tweak the plot appearance, so have
a look at the function documentation, `?plot2D`.

### Trench variance

The variance of the horizontal and vertical data variations can be assessed with

```{r}
estimateTrenchVariance(t13.trench2, .var = "d18O")
```

including confidence intervals based on the Chi-Squared distribution and the
number of effective degrees of freedom (DOF) in both trench directions. Default
for the latter are the number of data points in the trench directions; to
account for autocorrelation of the data the DOF can be supplied also manually
(see also `getEffectiveTrenchDOF` above).

### Estimating trench correlations and signal-to-noise ratios

An interesting application for a trench dataset is to study how the correlation
between neighbouring profiles varies with their separation distance. In
`TrenchR`, you can assess this question using `estimateInterProfileCorrelation`,
which calculates the average pairwise correlation between trench profiles for
given separation distances (including some tolerance so that the inter-profile
distances do not all need to exactly match the asked distances): 

```{r}
estimateInterProfileCorrelation(t13.trench2, distances = c(10, 30))
```

As you can see, the output gives the average correlation for each distance and
the associated standard deviation and standard error of the mean estimate. See
`?estimateInterProfileCorrelation` for details and more information on the error
estimation.

You can also calculate the correlation between profiles from two separate
trenches (assuming the two trenches were sampled at the same time):

```{r}
raw <- estimateInterTrenchCorrelation(t13.trench1, t13.trench2)
opt <- estimateInterTrenchCorrelation(t13.trench1, t13.trench2,
                                      optimize = TRUE, lag = -4 : 4)

str(raw)
str(opt)

mean(raw)
mean(opt$cor)
```

The first result is simply a matrix of the correlations for all possible pairs
of profiles between the two trenches. In the second example, some vertical
shifts of $\pm 4$ (depth) bins are allowed via the `lag` parameter in order to
optimise (i.e. maximise) the correlations; the result in this case is a list of
two matrices where the first matrix includes the optimised correlations between
the profiles and the second matrix the lags for which the optimal correlations
were observed.

Finally, you can estimate the signal-to-noise ratio of your trench dataset:

```{r}
estimateSNR(t13.trench2, distances = c(10, 30))
```

which is internally based on `estimateInterProfileCorrelation` and calculates
the signal-to-noise ratio (SNR) from the mean across the average pairwise
profile correlations evaluated at the given profile separation distances via
$\mathrm{SNR} = \overline{c}\,/\,(1 - \overline{c})$, where $\overline{c}$ is
the mean correlation.

## References

M端nch, T., Kipfstuhl, S., Freitag, J., Meyer, H., and Laepple, T.: Regional
climate signal vs. local noise: a two-dimensional view of water isotopes in
Antarctic firn at Kohnen Station, Dronning Maud Land, Clim. Past, 12(7),
1565-1581, doi:
[10.5194/cp-12-1565-2016](https://doi.org/10.5194/cp-12-1565-2016), 2016.

M端nch, T., Kipfstuhl, S., Freitag, J., Meyer, H., and Laepple, T.: Constraints
on post-depositional isotope modifications in East Antarctic firn from analysing
temporal changes of isotope profiles, The Cryosphere, 11(5), 2175-2188, doi:
[10.5194/tc-11-2175-2017](https://doi.org/10.5194/tc-11-2175-2017), 2017.